---
title: "Week 7 practical"
output: html_document
date: '2022-11-23'
---

# Introduction

## Analysing Spatial Autocorrelation with Moran’s I, LISA and friends

Spatial autocorrelation: a measure of similarity between nearby data

## Data download

```{r message=FALSE, warning=FALSE}
# libraries to read in
library(janitor)
library(sf)
library(tidyverse)
```


```{r message=FALSE, warning=FALSE}
london_wards <- st_read("statistical-gis-boundaries-london/ESRI/London_Ward_CityMerged.shp") %>%
  st_transform(., 27700)
```

```{r message=FALSE, warning=FALSE}
ward_data <- read_csv("https://data.london.gov.uk/download/ward-profiles-and-atlas/772d2d64-e8c6-46cb-86f9-e52b4c7851bc/ward-profiles-excel-version.csv", locale = locale(encoding = "latin1"), na = c("NA", "n/a")) %>% 
  clean_names()

london_wards <- london_wards %>% 
  left_join(ward_data, 
            by = c("GSS_CODE" = "new_code"))%>%
  dplyr::distinct(GSS_CODE, .keep_all = T)%>%
  dplyr::select(GSS_CODE, ward_name, average_gcse_capped_point_scores_2014)
```

```{r message=FALSE, warning=FALSE}
# check that it's in the right projection
st_crs(london_wards)
```

Always plot the data to check it looks reasonable (e.g. there are no points in other countries / the latitude and longitude are correct)

```{r message=FALSE, warning=FALSE}
library(tmap)

blue_plaques <- st_read("open-plaques-london-2018-04-08.geojson") %>%
  st_transform(.,27700)
```

```{r message=FALSE, warning=FALSE}
# make the map
tmap_mode("plot")
tm_shape(london_wards) +
  tm_polygons(col = NA, alpha = 0.5) +
tm_shape(blue_plaques) +
  tm_dots(col = "blue")
```

# Data Cleaning

need to remove the blue plaques that fall outside of London

```{r message=FALSE, warning=FALSE}
summary(blue_plaques)
```

```{r message=FALSE, warning=FALSE}
blue_plaques_london <- blue_plaques[london_wards,]

tm_shape(london_wards) +
  tm_polygons(col = NA, alpha = 0.5) +
tm_shape(blue_plaques_london) +
  tm_dots(col = "blue")
```

# Data manipulation

measures of spatial autocorrelation: require continuous observations to be spatially referenced (i.e. attached to a spatial unit)

creating a continuous observation from the blue plaques data - count all of the blue plaques that fall within each ward using the st_join function
- st_intersects: for every geometry pair whether they intersect (dense matrix), or which elements intersect (sparse)

```{r message=FALSE, warning=FALSE}
points_sf_joined <- london_wards %>%
  mutate(n= lengths(st_intersects(.,blue_plaques_london)))%>%
  janitor::clean_names()%>%
  #calculate area
  mutate(area=st_area(.))%>%
  #then density of the points per ward
  mutate(density=n/area)%>%
  dplyr::select(density, ward_name, gss_code, n, average_gcse_capped_point_scores_2014)
```

create a choropleth map

```{r message=FALSE, warning=FALSE}
tm_shape(points_sf_joined) +
    tm_polygons("density",
        style="jenks",
        palette="PuOr",
        midpoint=NA,
        popup.vars=c("wardname", "density"),
        title="Blue Plaque Density")
```

# Weights Matrix

define a spatial weights matrix

```{r message=FALSE, warning=FALSE}
# read in libraries required
library(spdep)
```

```{r message=FALSE, warning=FALSE}
# calculate the centroids of all Wards in London

coordsW <- points_sf_joined%>%
  st_centroid()%>%
  st_geometry()
  
plot(coordsW, axes=TRUE)
```

spatial weight matrix represents the spatial element of our data, this means we are trying to conceptualize and model how parts of the data are linked (or not linked) to each other spatially, using rules that we will set

based on contiguity
- rook: common boundaries 
- queen: any point in common (corners and boundaries)

based on distance
- k defines the number of neighbours considered

```{r}
# simple binary matrix of queen's case neighbours (shared corners and edges)

# create a neighbours list
# queen=T - indicates queen neighbours
LWard_nb <- points_sf_joined %>%
  poly2nb(., queen=T)
```

```{r}
summary(LWard_nb)
```

```{r message=FALSE, warning=FALSE}
# plot them
plot(LWard_nb, st_geometry(coordsW), col="red")
# add a map underneath
plot(points_sf_joined$geometry, add=T)
```

## Matrix style

make a matrix the size of the number of neighbourhoods with values indicating if the elements in the rows are a neighbour or not

B: the basic binary coding (1/0)
W: row standardised (sums over all links to n)
C: globally standardised (sums over all links to n)
U: equal to C divided by the number of neighbours (sums over all links to unity)
S: the variance-stabilizing coding scheme proposed by Tiefelsdorf et al. 1999, p. 167-168 (sums over all links to n)

```{r message=FALSE, warning=FALSE}
# create a spatial weights matrix from these weights based on binary coding
Lward.lw <- LWard_nb %>%
  nb2mat(., style="B")

sum(Lward.lw)
```
indicates there are 3680 neighbours in total 
- with global standardisation: divide the 625 wards by the total number of neighbours meaning each spatial weight has a value of 0.169
- with row standardisation (1 is divided by the sum of the number of neighbours in each row): e.g. row 1 here sums to 6, meaning each weight would be 0.166 in row 1 only.

# Autocorrelation

```{r}
# create list type object for the calculation of Moran's I
# global standardisation
Lward.lw <- LWard_nb %>%
  nb2listw(., style="C")
```

## Moran's I

- whether we have clustered values (close to 1) or dispersed values (close to -1)
- calculate for the densities rather than raw values

```{r message=FALSE, warning=FALSE}
I_LWard_Global_Density <- points_sf_joined %>%
  pull(density) %>%
  as.vector()%>%
  moran.test(., Lward.lw)

I_LWard_Global_Density
```

## Geary's C

- whether similar values or dissimilar values are clustering
- falls between 0 and 2
  =1: no spatial autocorrelation
  <1: positive spatial autocorrelation or similar values clustering
  >1 - negative spatial autocorreation or dissimilar values clustering)

```{r message=FALSE, warning=FALSE}
C_LWard_Global_Density <- 
  points_sf_joined %>%
  pull(density) %>%
  as.vector()%>%
  geary.test(., Lward.lw)

C_LWard_Global_Density
```

## Getis ord
- whether high or low values are clustering
  G > Expected = High values clustering
  G < expected = low values clustering

```{r}
G_LWard_Global_Density <- 
  points_sf_joined %>%
  pull(density) %>%
  as.vector()%>%
  globalG.test(., Lward.lw)

G_LWard_Global_Density
```

## Summary

The Moran’s I statistic = 0.67 (remember 1 = clustered, 0 = no pattern, -1 = dispersed) which shows that we have some distinctive clustering

The Geary’s C statistic = 0.41 which shows that similar values are clustering

The General G statistic = G > expected, so high values are tending to cluster.

```{r}
# Calculation of local Moran's I 
# use the localmoran function to generate I for each ward in the city

I_LWard_Local_count <- points_sf_joined %>%
  pull(n) %>%
  as.vector()%>%
  localmoran(., Lward.lw)%>%
  as_tibble()

I_LWard_Local_Density <- points_sf_joined %>%
  pull(density) %>%
  as.vector()%>%
  localmoran(., Lward.lw)%>%
  as_tibble()

# what does the output (the localMoran object) look like?
slice_head(I_LWard_Local_Density, n=5)
```

```{r message=FALSE, warning=FALSE}
# add the I score and the z score standard deviation 
points_sf_joined <- points_sf_joined %>%
  mutate(plaque_count_I = as.numeric(I_LWard_Local_count$Ii))%>%
  mutate(plaque_count_Iz =as.numeric(I_LWard_Local_count$Z.Ii))%>%
  mutate(density_I =as.numeric(I_LWard_Local_Density$Ii))%>%
  mutate(density_Iz =as.numeric(I_LWard_Local_Density$Z.Ii))
```


# Mapping

mapping the local Moran's I output
- whether the values are statistically different from the mean based on standard deviations at the 99% significance level (<-2.85 & >2.85), 95% significance level(<-1.96 & >1.96), and 90%(<-1.65% & >1.65)

```{r message=FALSE, warning=FALSE}
# create the breaks
breaks1<-c(-1000,-2.58,-1.96,-1.65,1.65,1.96,2.58,1000)
```

```{r}
# create a diverging colour brewing palette 
# reverse the order so higher values correspond to red
library(RColorBrewer)
MoranColours<- rev(brewer.pal(8, "RdGy"))
```

```{r message=FALSE, warning=FALSE}
tm_shape(points_sf_joined) +
    tm_polygons("plaque_count_Iz",
        style="fixed",
        breaks=breaks1,
        palette=MoranColours,
        midpoint=NA,
        title="Local Moran's I, Blue Plaques in London")
```

mapping the local Getis-ord output
- indication of hot spots and cold spots

```{r message=FALSE, warning=FALSE}
# Calculation of local Getis ord
Gi_LWard_Local_Density <- points_sf_joined %>%
  pull(density) %>%
  as.vector()%>%
  localG(., Lward.lw)

# only contains the z-score
head(Gi_LWard_Local_Density)

```

```{r}
points_sf_joined <- points_sf_joined %>%
  mutate(density_G = as.numeric(Gi_LWard_Local_Density))
```


```{r}
library(RColorBrewer)

GIColours<- rev(brewer.pal(8, "RdBu"))

#now plot on an interactive map
tm_shape(points_sf_joined) +
    tm_polygons("density_G",
        style="fixed",
        breaks=breaks1,
        palette=GIColours,
        midpoint=NA,
        title="Gi*, Blue Plaques in London")
```

# Other variables

Average GSCE scores

```{r message=FALSE, warning=FALSE}
# use head to see what other variables are in the data file
slice_head(points_sf_joined, n=2)
```

```{r message=FALSE, warning=FALSE}
# calculate the local Moran's I
I_LWard_Local_GCSE <- london_wards %>%
  arrange(GSS_CODE)%>%
  pull(average_gcse_capped_point_scores_2014) %>%
  as.vector()%>%
  localmoran(., Lward.lw)%>%
  as_tibble()

# add the column to the original dataframe
points_sf_joined <- points_sf_joined %>%
  arrange(gss_code)%>%
  mutate(GCSE_LocIz = as.numeric(I_LWard_Local_GCSE$Z.Ii))

# create the map
tm_shape(points_sf_joined) +
    tm_polygons("GCSE_LocIz",
        style="fixed",
        breaks=breaks1,
        palette=MoranColours,
        midpoint=NA,
        title="Local Moran's I, GCSE Scores")
```

```{r}
# calculate the local Getis Ord for average GCSE score
G_LWard_Local_GCSE <- points_sf_joined %>%
  dplyr::arrange(gss_code)%>%
  dplyr::pull(average_gcse_capped_point_scores_2014) %>%
  as.vector()%>%
  localG(., Lward.lw)

# add the column to the original dataframe
points_sf_joined <- points_sf_joined %>%
  dplyr::arrange(gss_code)%>%
  dplyr::mutate(GCSE_LocGiz = as.numeric(G_LWard_Local_GCSE))

# create the map
tm_shape(points_sf_joined) +
    tm_polygons("GCSE_LocGiz",
        style="fixed",
        breaks=breaks1,
        palette=GIColours,
        midpoint=NA,
        title="Gi*, GCSE Scores")
```


